---
title: "Do 311 Rat-Related Complaints Predict Initial Inspection Activity in NYC's
  Rat Mitigation Zones?"
author: "Siva Selvam"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: '3'
subtitle: VTPEH 6270 — Final Report
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.pos   = "H",
  out.width = "90%"
)
```

```{r packages}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(broom)
library(knitr)
library(tidyr)
library(purrr)
```

```{r load-data}
# Load all four datasets using relative paths (repo root)
complaints_raw <- read_csv("data/311-complaints.csv", show_col_types = FALSE) %>%
  rename(date = Date, zone_id = zoneID, rmz = RMZ,
         complaint_type = ComplaintType, count = Number) %>%
  mutate(date = mdy(date))

initial_raw <- read_csv("data/initial-inspections.csv", show_col_types = FALSE) %>%
  rename(date = Date, zone_id = zoneID, rmz = RMZ,
         inspection_type = Type, count = Number) %>%
  mutate(date = mdy(date))

compliance_raw <- read_csv("data/compliance-inspections.csv", show_col_types = FALSE) %>%
  rename(date = Date, zone_id = zoneID, rmz = RMZ,
         inspection_type = Type, count = Number) %>%
  mutate(date = mdy(date))

exterminator_raw <- read_csv("data/exterminator-visits.csv", show_col_types = FALSE) %>%
  rename(date = Date, zone_id = zoneID, rmz = RMZ,
         visit_type = Thing, count = Number) %>%
  mutate(date = mdy(date))

# Target RMZ
TARGET_RMZ <- "Manhattan: East Village and Chinatown"

# Build analysis dataset: semiannual totals for target RMZ
complaints_rmz <- complaints_raw %>%
  filter(rmz == TARGET_RMZ) %>%
  group_by(date) %>%
  summarise(complaints_n = sum(count, na.rm = TRUE), .groups = "drop")

initial_rmz <- initial_raw %>%
  filter(rmz == TARGET_RMZ) %>%
  group_by(date) %>%
  summarise(initial_inspections_n = sum(count, na.rm = TRUE), .groups = "drop")

analysis_df <- complaints_rmz %>%
  inner_join(initial_rmz, by = "date") %>%
  arrange(date) %>%
  mutate(
    period_label    = format(date, "%b %Y"),
    complaint_burden = if_else(
      complaints_n > median(complaints_n, na.rm = TRUE),
      "Higher complaint period", "Lower complaint period"),
    complaint_burden = factor(complaint_burden,
      levels = c("Lower complaint period", "Higher complaint period"))
  )

# Fit linear model
fit <- lm(initial_inspections_n ~ complaints_n, data = analysis_df)
cor_result <- cor.test(analysis_df$complaints_n,
                       analysis_df$initial_inspections_n,
                       method = "pearson")

# Observed model parameters
OBS_SLOPE     <- round(coef(fit)[2], 4)
OBS_INTERCEPT <- round(coef(fit)[1], 4)
OBS_NOISE     <- round(sigma(fit), 4)
OBS_R2        <- round(summary(fit)$r.squared, 3)
OBS_R         <- round(cor_result$estimate, 3)
OBS_P         <- round(cor_result$p.value, 4)
```

```{r theme}
report_theme <- function(base = 11) {
  theme_classic(base_size = base) +
    theme(
      plot.title    = element_text(face = "bold", size = base + 1),
      plot.subtitle = element_text(color = "grey40", size = base - 1),
      plot.caption  = element_text(color = "grey50", size = base - 2,
                                    hjust = 0, margin = margin(t = 6)),
      axis.title    = element_text(face = "bold"),
      legend.position  = "top",
      legend.title     = element_text(face = "bold"),
      panel.grid.major.y = element_line(linewidth = 0.25, color = "grey88"),
      strip.background   = element_rect(fill = "#f0f0f0", color = NA),
      strip.text         = element_text(face = "bold")
    )
}

# Color-blind friendly palette (Okabe-Ito)
rmz_colors <- c(
  "Brooklyn: Bed Stuy and Bushwick"       = "#0072B2",
  "Bronx: Grand Concourse"                = "#D55E00",
  "Manhattan: East Village and Chinatown" = "#E69F00",
  "Manhattan: Harlem"                     = "#009E73"
)

burden_colors <- c(
  "Lower complaint period"  = "#56B4E9",
  "Higher complaint period" = "#D55E00"
)
```

\newpage

# Introduction

## Background

Urban rat populations represent a persistent public health concern in high-density cities. In New York City, rats are associated with property damage, food contamination, and the potential transmission of zoonotic pathogens [@Himsworth2013]. In response, the New York City Department of Health and Mental Hygiene (NYC DOHMH) established the Rat Mitigation Zone (RMZ) program, which concentrates inspection and enforcement resources in neighborhoods with elevated rat activity [@NYCDOHMH2024].

Public complaints submitted via the city's 311 system offer one mechanism through which rat activity comes to the attention of municipal agencies. Prior research suggests that residents report rat sightings when they expect a visible municipal response [@Brown2017], and that shifts in human activity — such as those observed during the COVID-19 pandemic — can measurably alter both rat behavior and municipal response patterns [@Parsons2021]. Examining whether complaint volumes correspond to inspection activity can clarify the nature of this relationship: reactive (complaints drive inspections), proactive (both rise together during high-activity periods), or both.

## Research Question

**Is the number of initial inspections higher during semiannual periods with more 311 rat-related complaints in the Manhattan: East Village and Chinatown Rat Mitigation Zone?**

## Hypothesis

Semiannual periods with higher 311 complaint counts will also show higher numbers of initial inspections, consistent with a positive linear association between the two variables.

## Links

- **GitHub Repository:** [https://github.com/cornell-mba/vtpeh6270-rat-rmz](https://github.com/cornell-mba/vtpeh6270-rat-rmz)
- **Interactive Shiny App:** [https://s0mzwx-siva-selvam.shinyapps.io/shiny_app/](https://s0mzwx-siva-selvam.shinyapps.io/shiny_app/)

\newpage

# Material and Methods

## Data

Data were obtained from the NYC DOHMH as part of the Rat Mitigation Zone initiative [@NYCDOHMH2024]. Four publicly available datasets were used:

```{r data-overview-table}
data_tbl <- data.frame(
  File = c("311-complaints.csv", "initial-inspections.csv",
           "compliance-inspections.csv", "exterminator-visits.csv"),
  Description = c(
    "Semiannual 311 rat-related complaint counts by type and RMZ",
    "Semiannual initial inspection counts by outcome and RMZ",
    "Semiannual compliance inspection counts by outcome and RMZ",
    "Semiannual exterminator visit and bait application counts by RMZ"
  ),
  Rows = c(nrow(complaints_raw), nrow(initial_raw),
            nrow(compliance_raw), nrow(exterminator_raw))
)
kable(data_tbl, booktabs = TRUE,
      caption = "Overview of the four NYC DOHMH Rat Mitigation Zone datasets used in this analysis.")
```

All datasets cover **June 2022 through June 2025** in semiannual reporting periods (ending June 30 and December 31). Four RMZs are represented: Brooklyn: Bed Stuy and Bushwick, Bronx: Grand Concourse, Manhattan: East Village and Chinatown, and Manhattan: Harlem.

**This analysis is restricted to the Manhattan: East Village and Chinatown RMZ**, which provides seven semiannual observations (n = 7). No missing values were present after loading and cleaning.

## Variables

```{r variable-table}
var_tbl <- data.frame(
  Variable = c("complaints\\_n", "initial\\_inspections\\_n",
               "date", "period\\_label", "complaint\\_burden"),
  Type = c("Discrete count", "Discrete count", "Discrete (time)",
           "Character", "Ordered factor"),
  Description = c(
    "Total 311 complaints summed across all complaint types (semiannual)",
    "Total initial inspections summed across all inspection types (semiannual)",
    "End date of the semiannual reporting period (e.g., 2022-06-30)",
    "Human-readable period label (e.g., Jun 2022)",
    "Binary: Lower vs. Higher complaint period (split at median)"
  )
)
kable(var_tbl, booktabs = TRUE,
      caption = "Key variables used in the primary analysis.")
```

The primary exposure variable is **complaints\\_n** (Variable A) and the primary outcome variable is **initial\\_inspections\\_n** (Variable B). Both are analyzed on their original numeric scale; no mathematical transformations were applied.

## Statistical Analyses

The following analytical steps were conducted:

**1. Descriptive statistics** — means, standard deviations, and ranges were computed for both variables across the seven semiannual periods.

**2. Assumption checks** — prior to the primary analysis:

  - *Normality*: Shapiro-Wilk test applied to both variables. Given n = 7, this test has limited statistical power; results are interpreted alongside visual inspection of the scatter plot.
  - *Linearity*: assessed visually using a scatter plot with a fitted regression line.

**3. Pearson product-moment correlation** — used as the primary statistical test to quantify the direction and strength of the linear association between complaints and initial inspections. The significance threshold was set at $\alpha = 0.05$.

**4. Simple linear regression** — fit to estimate the magnitude of the association (slope) and model fit (R²):

$$\text{Initial Inspections} = \beta_0 + \beta_1 \times \text{Complaints} + \varepsilon$$

**5. Simulation study** — to assess how effect size (slope), sample size, and noise influence the detectability of the association. Parameters were drawn from the observed regression model, and 300 simulations were run across a grid of effect sizes, sample sizes, and noise levels.

All analyses were conducted in R (version 4.5.2). All code is available at the GitHub repository linked above.

\newpage

# Results

## Descriptive Summary

```{r desc-table}
desc_tbl <- analysis_df %>%
  summarise(
    `Mean complaints`       = round(mean(complaints_n), 1),
    `SD complaints`         = round(sd(complaints_n), 1),
    `Range complaints`      = paste0(min(complaints_n), " – ", max(complaints_n)),
    `Mean initial insp.`    = round(mean(initial_inspections_n), 1),
    `SD initial insp.`      = round(sd(initial_inspections_n), 1),
    `Range initial insp.`   = paste0(min(initial_inspections_n), " – ",
                                      max(initial_inspections_n)),
    N = n()
  )
kable(desc_tbl, booktabs = TRUE,
      caption = paste("Descriptive statistics for 311 complaints and initial inspections",
                      "(n = 7 semiannual periods, Manhattan: East Village and Chinatown RMZ)."))
```

Over the study period, mean semiannual 311 complaints were `r round(mean(analysis_df$complaints_n), 1)` (SD = `r round(sd(analysis_df$complaints_n), 1)`), ranging from `r min(analysis_df$complaints_n)` to `r max(analysis_df$complaints_n)`. Mean initial inspections were `r round(mean(analysis_df$initial_inspections_n), 1)` (SD = `r round(sd(analysis_df$initial_inspections_n), 1)`), ranging from `r min(analysis_df$initial_inspections_n)` to `r max(analysis_df$initial_inspections_n)`.

## Trends Over Time

```{r fig-trends, fig.height=4.2, fig.cap="Semiannual trends in 311 rat-related complaints (top) and initial inspections (bottom) for the Manhattan: East Village and Chinatown RMZ, June 2022 – June 2025. Both variables show a declining trend over the study period. Points represent observed semiannual totals; lines connect consecutive periods."}

complaints_plot <- analysis_df %>%
  ggplot(aes(x = date, y = complaints_n)) +
  geom_line(color = "#E69F00", linewidth = 1) +
  geom_point(color = "#E69F00", size = 3) +
  geom_text(aes(label = period_label), vjust = -0.85, size = 2.8, color = "grey40") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "6 months") +
  scale_y_continuous(labels = comma, limits = c(250, 450)) +
  labs(x = NULL, y = "311 Complaints\n(semiannual total)") +
  report_theme() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

inspections_plot <- analysis_df %>%
  ggplot(aes(x = date, y = initial_inspections_n)) +
  geom_line(color = "#0072B2", linewidth = 1) +
  geom_point(color = "#0072B2", size = 3) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "6 months") +
  scale_y_continuous(labels = comma, limits = c(6800, 9500)) +
  labs(x = "Reporting Period", y = "Initial Inspections\n(semiannual total)") +
  report_theme() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

gridExtra::grid.arrange(complaints_plot, inspections_plot, ncol = 1)
```

Both 311 complaints and initial inspections show a declining pattern over the study period (Figure 1). Complaints peaked at `r max(analysis_df$complaints_n)` in `r analysis_df$period_label[which.max(analysis_df$complaints_n)]` and reached a low of `r min(analysis_df$complaints_n)` in `r analysis_df$period_label[which.min(analysis_df$complaints_n)]`. Initial inspections followed a similar trajectory, peaking at `r max(analysis_df$initial_inspections_n)` and declining to `r min(analysis_df$initial_inspections_n)` by `r analysis_df$period_label[which.min(analysis_df$initial_inspections_n)]`.

## Assumption Checks

### Normality

```{r shapiro-table}
sw_c <- shapiro.test(analysis_df$complaints_n)
sw_i <- shapiro.test(analysis_df$initial_inspections_n)

sw_tbl <- data.frame(
  Variable = c("311 complaints", "Initial inspections"),
  W        = round(c(sw_c$statistic, sw_i$statistic), 4),
  `p-value` = round(c(sw_c$p.value, sw_i$p.value), 4),
  Conclusion = c("No evidence against normality",
                 "No evidence against normality"),
  check.names = FALSE
)
kable(sw_tbl, booktabs = TRUE,
      caption = "Shapiro-Wilk normality test results for both analysis variables (n = 7).")
```

Both variables returned Shapiro-Wilk p-values greater than 0.05 (Table 3), providing no evidence against the normality assumption. Given the small sample size (n = 7), this test has limited statistical power and results should be interpreted alongside the scatter plot visual.

### Linearity

```{r fig-linearity, fig.height=3.6, fig.cap="Scatter plot of 311 complaints versus initial inspections with a fitted linear regression line (black) and 95\\% confidence band (grey). The positive slope and reasonable point alignment support the linearity assumption. Manhattan: East Village and Chinatown RMZ, June 2022 – June 2025."}

ggplot(analysis_df, aes(x = complaints_n, y = initial_inspections_n)) +
  geom_smooth(method = "lm", se = TRUE, color = "black",
              fill = "grey80", linewidth = 0.9) +
  geom_point(size = 3, color = "#0072B2", alpha = 0.9) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    x = "311 rat-related complaints (semiannual total)",
    y = "Initial inspections (semiannual total)",
    caption = "Shaded band = 95% confidence interval of the regression line."
  ) +
  report_theme()
```

The scatter plot (Figure 2) shows a positive linear trend with points following the regression line reasonably well, supporting the linearity assumption.

## Primary Visualization

```{r fig-labeled-scatter, fig.height=4.2, fig.cap="Association between semiannual 311 rat-related complaints and initial inspections. Each point represents one semiannual reporting period, labeled by month and year. The red regression line with 95\\% confidence band illustrates the positive linear association. Points are colored by complaint burden category (split at median). Manhattan: East Village and Chinatown RMZ, June 2022 – June 2025."}

ggplot(analysis_df,
       aes(x = complaints_n, y = initial_inspections_n,
           color = complaint_burden, label = period_label)) +
  geom_smooth(method = "lm", se = TRUE,
              color = "#CC79A7", fill = "#f5c6d0",
              linewidth = 0.9, show.legend = FALSE) +
  geom_point(size = 3.5) +
  geom_text(vjust = -0.85, hjust = 0.5, size = 2.9, color = "grey35",
            show.legend = FALSE) +
  scale_color_manual(values = burden_colors, name = "Complaint burden") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma, limits = c(6800, 9700)) +
  labs(
    title = "Association Between 311 Complaints and Initial Inspections",
    subtitle = "Manhattan: East Village and Chinatown RMZ (2022–2025)",
    x = "311 rat-related complaints (semiannual total)",
    y = "Initial inspections (semiannual total)",
    caption = "Shaded band = 95% confidence interval of the regression line.\nPoints colored by complaint burden (split at median complaints_n)."
  ) +
  report_theme()
```

Figure 3 shows a consistent positive pattern: periods with the highest complaints (Jun 2022, Dec 2022) correspond to the highest inspection counts, while periods with the fewest complaints (Dec 2024, Jun 2025) correspond to the lowest inspection counts.

## Grouped Summary by Complaint Burden

```{r grouped-summary}
summary_tbl <- analysis_df %>%
  group_by(`Complaint burden` = complaint_burden) %>%
  summarise(
    N = n(),
    `Mean inspections` = round(mean(initial_inspections_n), 0),
    `SD inspections`   = round(sd(initial_inspections_n), 0),
    `Median inspections` = round(median(initial_inspections_n), 0),
    `Mean complaints`  = round(mean(complaints_n), 0),
    .groups = "drop"
  )
kable(summary_tbl, booktabs = TRUE,
      caption = paste("Initial inspections by semiannual complaint burden",
                      "(Manhattan: East Village and Chinatown RMZ)."))
```

Higher-complaint periods (n = 3) showed a mean of `r summary_tbl$"Mean inspections"[2]` initial inspections compared to `r summary_tbl$"Mean inspections"[1]` in lower-complaint periods (n = 4), a difference of approximately `r summary_tbl$"Mean inspections"[2] - summary_tbl$"Mean inspections"[1]` inspections.

## Statistical Results

### Pearson Correlation

```{r cor-table}
cor_tbl <- data.frame(
  `Pearson r`     = round(cor_result$estimate, 4),
  `t statistic`   = round(cor_result$statistic, 4),
  `df`            = cor_result$parameter,
  `p-value`       = round(cor_result$p.value, 4),
  `95% CI lower`  = round(cor_result$conf.int[1], 4),
  `95% CI upper`  = round(cor_result$conf.int[2], 4),
  check.names     = FALSE
)
kable(cor_tbl, booktabs = TRUE,
      caption = paste("Pearson correlation between 311 complaints and initial",
                      "inspections (n = 7, Manhattan: East Village and Chinatown RMZ)."))
```

The Pearson correlation coefficient was **r = `r OBS_R`** (t(5) = `r round(cor_result$statistic, 3)`, p = `r OBS_P`, 95% CI: [`r round(cor_result$conf.int[1], 3)`, `r round(cor_result$conf.int[2], 3)`]). This indicates a **strong positive linear association** between 311 complaints and initial inspections, statistically significant at the $\alpha = 0.05$ level.

### Simple Linear Regression

```{r reg-table}
reg_tbl <- tidy(fit) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  rename(Term = term, Estimate = estimate,
         `Std. Error` = std.error,
         `t value` = statistic, `p-value` = p.value) %>%
  mutate(Term = recode(Term,
    "(Intercept)" = "Intercept (β₀)",
    "complaints_n" = "311 Complaints (β₁)"))

kable(reg_tbl, booktabs = TRUE,
      caption = paste("Simple linear regression: initial inspections predicted by",
                      "311 complaints (n = 7, Manhattan: East Village and Chinatown RMZ)."))
```

```{r fit-stats}
fit_tbl <- data.frame(
  `R-squared`     = OBS_R2,
  `Adj. R-squared`= round(summary(fit)$adj.r.squared, 3),
  `F statistic`   = round(summary(fit)$fstatistic[1], 3),
  `Residual SE`   = round(sigma(fit), 1),
  check.names     = FALSE
)
kable(fit_tbl, booktabs = TRUE,
      caption = "Model fit statistics for simple linear regression.")
```

The regression model estimates that **each additional 311 complaint is associated with approximately `r round(OBS_SLOPE, 1)` additional initial inspections** (SE = `r round(tidy(fit)$std.error[2], 2)`, p = `r round(tidy(fit)$p.value[2], 4)`). The model explains **`r OBS_R2 * 100`% of the variance** in initial inspections (R² = `r OBS_R2`).

## Simulation Study

To contextualize detectability of the observed association, a simulation study explored how effect size (slope), sample size (n), and noise (residual SD) influence the inspection gap between the lowest and highest complaint quartiles.

```{r simulation, cache=FALSE}
set.seed(42)

# Observed parameters
obs_slope  <- OBS_SLOPE
obs_int    <- OBS_INTERCEPT
obs_noise  <- OBS_NOISE
comp_min   <- min(analysis_df$complaints_n)
comp_max   <- max(analysis_df$complaints_n)

# Parameter grid (condensed for report)
effect_sizes <- seq(max(0, obs_slope * 0.2), obs_slope * 1.8, length.out = 8)
sample_sizes <- c(7, 10, 20, 50, 100)
noise_levels <- c(obs_noise * 0.5, obs_noise, obs_noise * 1.5)

param_grid <- expand.grid(slope = effect_sizes,
                           n     = sample_sizes,
                           noise = noise_levels)

simulate_one <- function(slope, n, noise) {
  comp_sim <- runif(n, comp_min, comp_max)
  insp_sim <- obs_int + slope * comp_sim + rnorm(n, 0, noise)
  insp_sim <- pmax(0, insp_sim)
  data.frame(complaints_n = comp_sim, inspections = insp_sim) %>%
    mutate(q = ntile(complaints_n, 4)) %>%
    group_by(q) %>%
    summarise(mean_insp = mean(inspections), .groups = "drop") %>%
    summarise(gap = mean_insp[q == 4] - mean_insp[q == 1])
}

sim_results <- pmap_dfr(param_grid, function(slope, n, noise) {
  simulate_one(slope, n, noise) %>%
    mutate(slope = slope, n = n, noise = noise)
})
```

```{r fig-simulation, fig.height=4.5, fig.cap="Simulated inspection gap (Q4 minus Q1 mean initial inspections) across effect sizes and sample sizes, faceted by noise level. Warmer colors indicate a larger detectable gap between the highest and lowest complaint quartiles. The observed slope (≈10.4) and noise level (≈440) are marked by vertical and horizontal dashed lines respectively. Results demonstrate that the observed positive association is reliably detectable at moderate-to-large sample sizes and the observed effect size."}

obs_slope_line <- obs_slope

sim_results %>%
  mutate(noise_label = paste0("Noise SD = ", round(noise, 0))) %>%
  ggplot(aes(x = factor(n), y = factor(round(slope, 2)), fill = gap)) +
  geom_tile(color = "white", linewidth = 0.4) +
  facet_wrap(~ noise_label, ncol = 3) +
  scale_fill_gradient2(
    low      = "#0072B2",
    mid      = "#FFFFBF",
    high     = "#D55E00",
    midpoint = 0,
    name     = "Inspection\nGap (Q4 − Q1)",
    labels   = comma
  ) +
  labs(
    x = "Sample Size (n)",
    y = "Effect Size (slope)",
    caption = paste0("Observed slope ≈ ", round(obs_slope, 1),
                     " | Observed noise SD ≈ ", round(obs_noise, 0),
                     " | Intercept fixed at observed value (", round(obs_int, 0), ").")
  ) +
  report_theme() +
  theme(
    axis.text.x  = element_text(angle = 35, hjust = 1),
    legend.position = "right",
    panel.grid   = element_blank()
  )
```

The simulation (Figure 5) confirms that at the observed slope (~`r round(obs_slope, 1)`) and noise level (~`r round(obs_noise, 0)`), the inspection gap between the lowest and highest complaint quartiles is detectable and positive. Higher noise and smaller sample sizes attenuate and destabilize the estimated gap, consistent with statistical theory. The observed dataset (n = 7) falls at the lower end of the sample size axis, underscoring the importance of interpreting results cautiously.

\newpage

# Conclusions

## Summary of Findings

This analysis examined whether semiannual 311 rat-related complaint counts are positively associated with initial inspection counts in the Manhattan: East Village and Chinatown RMZ over seven reporting periods from June 2022 through June 2025.

A **strong positive Pearson correlation** was found (r = `r OBS_R`, p = `r OBS_P`, 95% CI: [`r round(cor_result$conf.int[1], 3)`, `r round(cor_result$conf.int[2], 3)`]), and the simple linear regression model explained **`r OBS_R2 * 100`% of the variance** in initial inspections (R² = `r OBS_R2`). Each additional semiannual 311 complaint was associated with approximately `r round(OBS_SLOPE, 1)` additional initial inspections.

These findings are consistent with the study hypothesis and align with prior literature. The positive association may reflect both reactive municipal response to complaint-driven signals and proactive inspection activity that co-occurs during periods of elevated rat activity [@Brown2017; @Himsworth2013]. The parallel decline in both complaints and inspections after 2022 is consistent with the wind-down of pandemic-related shifts in urban rat behavior documented by Parsons et al. [-@Parsons2021].

## Statistical vs. Practical Significance

The result is statistically significant at $\alpha = 0.05$. In practical terms, a difference of ~140 complaints between the highest (422 in Dec 2022) and lowest (280 in Dec 2024) periods corresponds to an estimated difference of approximately 1,450 initial inspections — a meaningful operational difference in municipal surveillance activity. However, correlation does not establish causation, and the shared temporal trend (both variables declining from 2022 onward) may partly explain the observed association, independent of any direct complaint-to-inspection mechanism.

## Limitations

1. **Small sample size (n = 7):** Statistical power is low. The Shapiro-Wilk and Pearson tests have limited reliability at this sample size, and confidence intervals are wide.
2. **Temporal confounding:** Both variables declined over time, which could produce a spurious correlation driven by a shared trend rather than a direct causal relationship.
3. **Single RMZ:** Results are specific to the Manhattan: East Village and Chinatown RMZ and may not generalize to other zones with different neighborhood characteristics, population density, or municipal resource allocation.
4. **Aggregated data:** Semiannual totals mask within-period variation. Monthly or weekly data would allow more precise temporal analysis.

## Future Directions

Future work could extend this analysis to all four RMZs using a multi-level or panel data framework, incorporate monthly or weekly observations, and test for Granger causality to better distinguish reactive from proactive inspection patterns. Linking complaint and inspection data to neighborhood-level socioeconomic indicators could also reveal equity-related patterns in municipal response.

\newpage

# AI Use Disclosure Statement

Yes, AI-based tools were used in completing this report. Claude (Anthropic) was used to assist with structuring the R Markdown report, generating draft R code for data loading, statistical analyses, visualizations, simulation, and the Shiny application, and assisting with written interpretation. All code, analyses, interpretations, and final outputs were reviewed and validated by the author.

\newpage

# References

Brown, K., Lombard, A. M., & Aronson, M. F. J. (2017). Resident perceptions of urban wildlife and factors associated with complaint behavior. *Human Dimensions of Wildlife*, 22(4), 375–384. https://doi.org/10.1080/10871209.2017.1320343

Himsworth, C. G., Parsons, K. L., Jardine, C., & Patrick, D. M. (2013). Rats, cities, people, and pathogens: A systematic review and narrative synthesis of literature regarding the ecology of rat-associated zoonoses in urban centers. *Vector-Borne and Zoonotic Diseases*, 13(6), 349–359. https://doi.org/10.1089/vbz.2012.1195

New York City Department of Health and Mental Hygiene. (2024). *Rat Mitigation Zones data*. https://a816-dohbesp.nyc.gov/IndicatorPublic/data-features/rat-mitigation-zones/

Parsons, M. H., et al. (2021). Rats and the COVID-19 pandemic: Considering the influence of social distancing on a global commensal pest. *Urban Ecosystems*, 24, 1079–1088. https://doi.org/10.1007/s11252-021-01107-9
